R Packages to Aid in Handling Web Access Logs

### 2015-12-27 paper/paper.Rmd
0.2.0

The rise of the World Wide Web over the last few decades has made it dramatically easier to access and transfer data, and R \citep{RCore} boasts abundant functionality when it comes to taking data *from* the web. Base R itself has simple file downloading and page reading capabilities, through the `download.file` and `readLines` functions, and additional functionality is made available for handling web-accessible data through packages such as \CRANpkg{httr} \citep{httr}.

@@

As a statistically-oriented programming language, R is commonly used by these same researchers for data analysis, testing and reporting - but it lacks tools designed for the kinds of data sources and formats that are encountered when dealing with access logs. In this article we review the use cases for particular operations over web data, the limitations in base R when it comes to performing those operations, and a suite of R packages designed to overcome them: \CRANpkg{webreadr} \citep{webreadr}, designed for reading access logs in,
\CRANpkg{urltools} \citep{urltools}, for URL manipulation, \CRANpkg{iptools} \citep{iptools} for handling IP addresses, and \CRANpkg{rgeolocate} \citep{rgeolocate} for direct IP geolocation.

@@

To make reading access logs into R as easy as possible we created the \pkg{webreadr} package. This contains user-friendly equivalents to \code{read.table} for each type of log, detecting the fields that should appear, converting the timestamps into POSIX objects, and merging fields or splitting fields where necessary. The package contains four core functions, one for each form of log: \code{read\_clf} for the Common Log Format, \code{read\_combined} for the Combined Log Format, and \code{read\_squid} and \code{read\_aws} for Squid and AWS formats respectively. Each one abstracts away the complexity of specifying column names and formats, and instead allows a researcher to read a file in with a minimal amount of work.

@@

To solve this common problem in analysing request logs, the \pkg{urltools} package was created. This contains a function,
\code{url\_decode}, which decodes URLs - but relies on vectorised, compiled code. Benchmarking the two approaches against each other shows that the \pkg{urltools} implementation is approximately 60-70 times faster over large datasets. Again using \pkg{microbenchmark}, if we compare the vectorised decoding of 1,000,000 URLs with \pkg{urltools} against \code{URLdecode} in a \code{vapply} loop, we see a 60-70 times speed improvement:

@@

Based around the \emph{Boost ASIO} C++ library \citep{boostcite}, \pkg{iptools} is a package designed for this kind of IP manipulation (and more). Built around combined code, it is extremely fast (for that code that does not rely on internet connections) and boasts a range of features.

@@

With these concerns in mind, we wrote the \pkg{rgeolocate} package. Through \pkg{httr} this contains convenient, vectorised bindings
to various web services that provide geographic metadata about IP addresses. More importantly, using the \CRANpkg{Rcpp} package to integrate C++ and R, \pkg{rgeolocate} also features a direct, compiled binding to the MaxMind API. This means that local binary databases can also be queried, which is far faster and more robust than web-based equivalents and avoids the privacy concerns associated with transmitting users' IP addresses externally.

>>>>

The rise of the World Wide Web over the last few decades has made it dramatically easier to access and transfer data, and R \citep{RCore} boasts abundant functionality when it comes to taking data *from* the web. Base R itself has simple file downloading and page reading capabilities, through the \code{download.file} and \code{readLines} functions, and additional functionality is made available for handling web-accessible data through packages such as \CRANpkg{httr} \citep{httr}.

@@

As a statistically-oriented programming language, R is commonly used by these same researchers for data analysis, testing and reporting - but it lacks tools designed for the kinds of data sources and formats that are encountered when dealing with access logs. In this article we review the use cases for particular operations over web data, the limitations in base R when it comes to performing those operations, and a suite of R packages designed to overcome them: reading access logs in, manipulating URLs, manipulating IP addresses, and direct IP geolocation.

@@

To make reading access logs into R as easy as possible we created the \CRANpkg{webreadr} \citep{webreadr} package. This contains user-friendly equivalents to \code{read.table} for each type of log, detecting the fields that should appear, converting the timestamps into POSIX objects, and merging fields or splitting fields where necessary. The package contains four core functions, one for each form of log: \code{read\_clf} for the Common Log Format, \code{read\_combined} for the Combined Log Format, and \code{read\_squid} and \code{read\_aws} for Squid and AWS formats respectively. Each one abstracts away the complexity of specifying column names and formats, and instead allows a researcher to read a file in with a minimal amount of work.

@@

To solve this common problem in analysing request logs, the\CRANpkg{urltools} \citep{urltools} package was created. This contains a function,
\code{url\_decode}, which decodes URLs and relies on vectorised, compiled code to do so. Benchmarking the two approaches against each other shows that the \pkg{urltools} implementation is approximately 60-70 times faster over large datasets. Again using \pkg{microbenchmark}, if we compare the vectorised decoding of 1,000,000 URLs with \pkg{urltools} against \code{URLdecode} in a \code{vapply} loop, we see a 60-70 times speed improvement:

@@

Based around the \emph{Boost ASIO} C++ library \citep{boostcite}, \CRANpkg{iptools} \citep{iptools} is a package designed for this kind of IP manipulation (and more). Built around combined code, it is extremely fast (for that code that does not rely on internet connections) and boasts a range of features.

@@

With these concerns in mind, we wrote the \CRANpkg{rgeolocate} \citep{rgeolocate} package. Through \pkg{httr} this contains convenient, vectorised bindings
to various web services that provide geographic metadata about IP addresses. More importantly, using the \CRANpkg{Rcpp} package to integrate C++ and R, \pkg{rgeolocate} also features a direct, compiled binding to the MaxMind API. This means that local binary databases can also be queried, which is far faster and more robust than web-based equivalents and avoids the privacy concerns associated with transmitting users' IP addresses externally.

### 2015-12-27 paper/paper.tex
add address
0.2.0

The rise of the World Wide Web over the last few decades has made it
dramatically easier to access and transfer data, and R \citep{RCore}
boasts abundant functionality when it comes to taking data \emph{from}
the web. Base R itself has simple file downloading and page reading
capabilities, through the \texttt{download.file} and \texttt{readLines}
functions, and additional functionality is made available for handling
web-accessible data through packages such as \CRANpkg{httr}
\citep{httr}.

@@

\address{%
Oliver Keyes\\
Wikimedia Foundation\\
line 1\\ line 2\\
}
\href{mailto:author1@work}{\nolinkurl{author1@work}}

>>>>

The rise of the World Wide Web over the last few decades has made it
dramatically easier to access and transfer data, and R \citep{RCore}
boasts abundant functionality when it comes to taking data \emph{from}
the web. Base R itself has simple file downloading and page reading
capabilities, through the \code{download.file} and \code{readLines}
functions, and additional functionality is made available for handling
web-accessible data through packages such as \CRANpkg{httr}
\citep{httr}.

@@

\address{%
Oliver Keyes\\
Wikimedia Foundation\\
149 New Montgomery Street, 6th floor\\ San Francisco, CA, 94105, USA\\
}
\href{mailto:ironholds@gmail.com}{\nolinkurl{ironholds@gmail.com}}
